A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-26 paper.tex
Tweak text on Ne >> n violations

% JK: This paragraph is weak, and needs some work. Basically want to
% illustrate the mathematical literature which views ancestral
% graphs very much as *processes*.
% Jere: ASG section redrafted, gene transfer left as TODO (because I
% know essentially nothing about it).
In parallel with this tendency to decouple the term
Ancestral Recombination Graph from the stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes encoded via a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode desired
patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.

@@

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of \emph{events}. Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In order to distinguish this form of ARG from what we will be discussing
in later sections we refer to it as an ``event ARG``, or eARG for brevity.
An example of this form of an eARG
is given in Fig.~\ref{fig-arg-data-structure}B.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage (e.g., node \noderef{e}
in Fig.~\ref{fig-arg-data-structure}), and in a recombination
event a lineage is split into two independent
ancestral lineages (e.g., node \noderef{d}
in Fig.~\ref{fig-arg-data-structure}). The final vital
detail---which distinguishes an arbitrary graph with the required topological
properties from an ARG---is that we associate a breakpoint with each recombination
event. This information is sufficient to uniquely
define the local genealogical trees at every position along the genome,
which is the basic requirement for a data structure encoding
genome-wide genetic ancestry.

@@

% What do we mean by an eARG, formally?
To be concrete, we can formally define the
classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(This formulation is equivalent to having different ``types''
of node, with breakpoints are associated with recombination
nodes only).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
This encoding of an eARG is equivalent to the output
produced by programs such as
\texttt{ARGweaver}~\citep{rasmussen2014genome}
and \texttt{KwARG}~\citep{ignatieva2021kwarg}:
an example is given in Fig.~\ref{fig-arg-data-structure}A.
Note that the description above captures only the
graph topology: if we also wish to know the branch lengths we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes
rather than edges: we need to know from which parent the
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG (the most basic downstream
operation for an ARG data structure).
To recover the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event above a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course, but depending on the ordering of
otherwise indistinguishable objects is generally problematic. In practice,
several methods explicitly associate information with the outbound edges
of a recombination event
to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.

@@

% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$. This
assumption ensures that multiple events (i.e., simultaneous
coalescence and recombination) don't occur at the same
time, which is reflected in the eARG definitions.
This once-reasonable assumption has been overtaken by
the data we have available.
% This can be tightened up a bit, I think. We may need to drop a
% few references.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
Such datasets are also available for other species,
for example, \citet{RosFreixedes2022} have combined genome
and SNP array data to infer recombination of genomes for 440,610 pigs within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.

>>>>

In parallel with this tendency to decouple the term
Ancestral Recombination Graph from the stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes described as a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode desired
patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
% Could mention the Ancestral Gene Transfer graph here too, within
% something like "Similarly the Ancestral Gene Transfer Graph [cite]
% and [other ancestral graphs, if they exist?] model genealogical
% processes in bacteria as a graph.

@@

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of \emph{events}. Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In order to distinguish this form of ARG from what we will be discussing
in later sections we refer to it as an ``event ARG``, or eARG for brevity.
An example of an eARG
is given in Fig.~\ref{fig-arg-data-structure}B.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage (e.g., node \noderef{e}
in Fig.~\ref{fig-arg-data-structure}), and in a recombination
event a lineage is split into two independent
ancestral lineages (e.g., node \noderef{d}
in Fig.~\ref{fig-arg-data-structure}). The final vital
detail---which distinguishes an arbitrary graph with the required topological
properties from an eARG---is that we associate a breakpoint with each recombination
event. This information is sufficient to uniquely
define the local genealogical trees at every position along the genome,
which is the basic requirement for a data structure encoding
genome-wide genetic ancestry.

@@

% What do we mean by an eARG, formally?
To be concrete, we can formally define the
classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(This is equivalent to having different ``types''
of node, where breakpoints are associated with recombination
nodes only).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
% For each recombination node $u$ we must
% have exactly two edges in $e$ such that $u$ is the child.
This encoding of an eARG is equivalent to the output
of programs such as
\texttt{ARGweaver}~\citep{rasmussen2014genome}
and \texttt{KwARG}~\citep{ignatieva2021kwarg},
and an example is given in Fig.~\ref{fig-arg-data-structure}A.
Note that the description above captures only the
graph topology: if we also wish to know the branch lengths we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes:
we need to know from which parent the node inherited
the segments of genome to the left and right of the breakpoint.
The simplest way to break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG (the most basic downstream
operation for an ARG data structure).
To recover the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event above a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course, but depending on the ordering of
otherwise indistinguishable objects is generally problematic. In practice,
several methods explicitly associate information with the outbound edges
of a recombination event
to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.

@@

% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$. This
assumption ensures that multiple events (i.e., simultaneous
coalescence and recombination) don't occur at the same
time, which is reflected in the eARG definitions.
This once-reasonable assumption has been overtaken by
the data we have available.
% This can be tightened up a bit, I think. We may need to drop a
% few references.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
In another example, genome and SNP array data for 440,610 individuals within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
